\vskip-0.6in Trade Agreements in the Shadow of Lobbying\protect\footnote{Formerly circulated under the title ``Trade Agreements, Lobbying and Separation of Powers''}

### 2016-02-02 draft.tex
revise 'how it achieves' and contributions' parts of intro

@@

Consider first the case of political certainty, and assume that the executive is more pro-social than the legislature, as has been the case throughout the post-war period in the United States.\footnote{The standard assumption that the executive is less protectionist than the legislature is a special case of the finding that susceptibility to special interests generally declines with constituency size. One simple illustration from the realm of trade policy is the following: a legislator whose district has a large concentration of a particular industry does not take into account the impact of tariffs on the welfare of consumers in other districts, while the executive, whose constituency encompasses the whole country, will internalize these diffuse consumption effects. See Lohmann and O'Halloran (1994) for a detailed argument and Grossman and Helpman (2005) for a formal model that demonstrates the point for the legislature.\label{fn:ga_l_e}} For any trade-agreement tariff set by the executives, the lobby can calculate the effort level required to prevent the agreement from being ratified: it must shift the preferences of the median legislator just far enough so that he will choose the non-cooperative tariffs over the trade agreement tariffs. Because the lobby will pay this price so long as its benefit outweighs the cost, a trade war will occur unless the executive sets the trade-agreement tariff so that lobbying is not worthwhile.\footnote{As will be made clear below, non-ratification is equivalent to breaking a trade agreement once in place; likewise, the non-cooperative tariffs that result from failing to ratify the trade agreement are the same as the tariffs that would arise in trade war, so this model applies to both situations.}

Thus, any trade agreement under political certainty involves a tariff high enough to disengage the lobby completely; that is, even welfare-maximizing executives set high tariffs and induce zero contributions in equilibrium because they set policy in the shadow of a ratification decision influenced by lobbying. Careful modeling of the political process demonstrates that, while tariff-setting behavior is closely linked to deadweight-loss calculations, the empirical estimates of the responsiveness of protective measures to import penetration and the elasticity of import demand do not necessarily map directly onto the PFS parameter for welfare-mindedness.

@@

One contribution of this model, therefore, is the idea that trade agreements can act as a kind of domestic political commitment device. Rather than a time inconsistency such as that in Maggi and Rodr\'{i}guez-Clare (2007) where the welfare of the unitary policymaker is improved via commitment to a trade agreement by changing the investment decisions of firms, here we have an inconsistency between the preferences of the executive and legislative branches.\footnote{Indeed, depending on the modeling choices and parameters, the utility of the legislative branch taken as a whole may be lower under the trade agreement scenario.} Although the legislature has the final word both with and without the trade agreement, the executives are able to use commitment to the trade agreement to reduce lobbying effort, thus muting the legislature's protectionist bias and bringing the outcome more in line with their relatively liberal preferences.

>>>>

To see how this works, consider first the case of political certainty, and assume that the executive is more pro-social than the legislature, as has been the case throughout the post-war period in the United States.\footnote{The standard assumption that the executive is less protectionist than the legislature is a special case of the finding that susceptibility to special interests generally declines with constituency size. One simple illustration from the realm of trade policy is the following: a legislator whose district has a large concentration of a particular industry does not take into account the impact of tariffs on the welfare of consumers in other districts, while the executive, whose constituency encompasses the whole country, will internalize these diffuse consumption effects. See Lohmann and O'Halloran (1994) for a detailed argument and Grossman and Helpman (2005) for a formal model that demonstrates the point for the legislature.\label{fn:ga_l_e}} For any trade-agreement tariff set by the executives, the lobby can calculate the effort level required to prevent the agreement from being ratified: it must shift the preferences of the median legislator just far enough so that he will choose the non-cooperative tariffs over the trade agreement tariffs. Because the lobby will pay this price so long as its benefit outweighs the cost, a trade war will occur unless the executive sets the trade-agreement tariff so that lobbying is not worthwhile.\footnote{As will be made clear below, non-ratification is equivalent to breaking a trade agreement once in place; likewise, the non-cooperative tariffs that result from failing to ratify the trade agreement are the same as the tariffs that would arise in trade war, so this model applies to both situations.}

Thus, any trade agreement under political certainty involves a tariff high enough to disengage the lobby completely; that is, even welfare-maximizing executives set high tariffs and induce zero contributions in equilibrium because they set policy in the shadow of a ratification decision influenced by lobbying. Careful modeling of the political process demonstrates that \textbf{the relationship between deadweight loss calculations and the tariffs that are provided in a trade agreement may be mediated in important ways by the ratification constraint.}

One contribution of this model, therefore, is the idea that trade agreements can act as a kind of domestic political commitment device. Rather than a time inconsistency such as that in Maggi and Rodr\'{i}guez-Clare (2007) where the welfare of the unitary policymaker is improved via commitment to a trade agreement by changing the investment decisions of firms, here we have an inconsistency between the preferences of the executive and legislative branches.\footnote{Indeed, depending on the modeling choices and parameters, the utility of the legislative branch taken as a whole may be lower under the trade agreement scenario.} Although the legislature has the final word both with and without the trade agreement, the executives are able to use the trade agreement to reduce lobbying effort, thus muting the legislature's protectionist bias and bringing the outcome more in line with their relatively liberal preferences.

@@

@@
